---
title: "extract_hhlevel_data_and_join"
author: "Jon Minton; Martin Taulbut"
format: html
editor: visual
---

## Aim

The aim of this document is to develop a good workflow for extracting data from the household level surveys in the UKHLS and link them to individuals.

## Dependencies

Here are some expected dependencies

```{r}
library(here)
library(tidyverse)
library(haven)
devtools::load_all(here::here("R"))
```

## Locations and files

We start by specifying where the files of interest are kept, and producing a list of individual and household level files

```{r}
base_dir_location <- here("big_data/UKDA-6614-stata/stata/stata13_se/ukhls/")
```

```{r}
hh_files_only <- dir(
  base_dir_location, 
  pattern = "^[a-z]{1}_hhresp.dta", 
  full.names = TRUE
)
```

We also need a list of all individual level files, as to get the pid and hid variables to link

```{r}
ind_files_only <- dir(
  base_dir_location, 
  pattern = "^[a-z]{1}_indresp.dta",
  full.names = TRUE
)
```

From the hh level data we want to extract the following

-   `{w}_hidp`

-   `{w}_fihhmnet1_dv` : monthly net income

-   `{w}_ieqmoecd_dv`: equivocation factor

-   `{w}_nkids_dv`: The number of dependent children in the household

Where {w} indicates wave letter.

For now, these variables are hardcoded in the function `extract_eq_income_and_num_dependents`

```{r}

all_hh_data <- 
  lapply(hh_files_only, extract_eq_income_and_num_dependents) %>% 
  bind_rows()

all_hh_data

```

We want to do something similar with the individual level data

```{r}




all_pid_hid_lookups <- lapply(
  ind_files_only, 
  extract_pid_with_hid_and_wave
) %>% 
  bind_rows()

all_pid_hid_lookups
```

Now to check I can link the pidp to the hh variables extracted

```{r}
all_hh_data %>% 
  right_join(all_pid_hid_lookups) %>% 
  select(pidp, hidp, wave, everything())

```

## Conclusion

This has shown that the functions now moved to the package work in the intended way.

To avoid this notebook becoming too complex I think we should do actual analyses linking the above with other individual level data in a different notebook with an appropriate name
