---
title: "Education/Qualification and Health Effects Summary"
author: "Jon Minton"
execute:
  echo: false
  warning: false
  message: false
format: 
  pptx:
    code-fold: true
    df-print: kable
editor: visual
---

## Modelling the effect of health on labour market engagement

We use three types of approach:

1.  Using **reported long-term illness** as a binary variable
2.  Using **scores on SF-12**, mental and health sub-domain, as a continuous variable
3.  Using reported presence or absence of **specific clinically diagnosed medical conditions** as a series of binary variables

We will now consider these three approaches in turn

## General approach (as a reminder)

The general approach taken is:

1.  We decide on the variables to consider including as exposures
2.  We fit a series of models with and without exposures to see if including the exposures adds to the quality of the model (Using the AIC/BIC criteria)
    1.  If a model with the exposure is preferred we use this for running baseline-counterfactual scenarios
    2.  Even if a model with exposure does not outcompete the foundational (no-exposure) specification we *may* still run it if there are sound clinical/epidemiological reasons to do so
3.  We decide on a representative dataset to pass to the best exposure model to estimate the effects on economic (in)activity composition in the next wave under a do-nothing (**baseline**) scenario
    1.  This has usually been wave J of UKHLS
4.  We modify the baseline dataset to allow the estimated effects of the exposure on economic (in)activity to be estimated. This produces one or more **counterfactual** scenarios
5.  We compare the **baseline** and **counterfactual** scenarios' estimated economic (in)activity compositions in absolute and relative terms:
    1.  Absolute differences: how many more/fewer people are (say) unemployed in the counterfactual scenario than the baseline scenario?
    2.  Relative differences: how many % higher/lower are the numbers (say) unemployed in the counterfactual scenario than the baseline scenario?

## Modelling based on reported long-term illness

Long-term illness is a binary variable: present/absent

## Jointly modelling education changes and health changes

Let's define this in PICO structure:

-   **Population** Representative adults aged 25-60 years

-   **Intervention:**

    -   Education: Everyone up one level (up to degree)

    -   Health: 1 SD unit change, equal through MH and PH path

-   **Comparator:** No change in health or qualifications

-   **Outcome:** Change in absolute and relative distribution of persons in EI/EA states, focusing on change in EI: LT sick

## Step 1: Get data

```{r}

devtools::load_all(here::here('R'))

base_dir_location <- "big_data/UKDA-6614-stata/stata/stata13_se/ukhls"
library(tidyverse)
library(nnet)


varnames <-  c(
  "jbstat", "dvage", "sex", 'hiqual_dv', "sf12mcs_dv", "sf12pcs_dv"
  )

extract_what <- c(
  "labels", "values", "labels", "labels", "values", "values"
  )

ind_data <- get_ind_level_vars_for_selected_waves(
  varnames = varnames, vartypes = extract_what
)

ind_data_standardised <- 
  ind_data |> 
  mutate(
    qual_group = case_when(
      hiqual_dv %in% c("No qual", "No qualification") ~ "None",
      hiqual_dv %in% c("GCSE etc", "A level etc", "A-level etc", "Other qual", "Other qualification", "Other higher") ~ "Some", 
      hiqual_dv %in% c("Degree", "Other higher degree") ~ "Degree",
      TRUE ~ NA_character_
    )
  ) |> 
  mutate(
    qual_group = ordered(qual_group, levels = c("None", "Some", "Degree"))
  ) |> 
  rename(age = dvage) |> 
  mutate(across(c(age, sf12mcs_dv, sf12pcs_dv), function(x) ifelse(x < 0, NA, x))) %>%
    mutate(across(c(sf12mcs_dv, sf12pcs_dv), standardise_scores)) |> 
  filter(between(age, 25, 60)) %>%  #As highest qualification starting at 25 not 16 years of age  |> 
  filter(complete.cases(.))


```

## Modelling 

```{r}

mod_00 <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5),
    data = ind_data_standardised
  )

mod_hlth <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + sf12pcs_dv*sf12mcs_dv,
    data = ind_data_standardised
  )

mod_qual <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + qual_group,
    data = ind_data_standardised
  )

mod_full <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + sf12pcs_dv*sf12mcs_dv + qual_group,
    data = ind_data_standardised
  )
  
  
```

## Model comparison 

```{r}
aics <- AIC(mod_00, mod_hlth, mod_qual, mod_full)$AIC
bics <- BIC(mod_00, mod_hlth, mod_qual, mod_full)$BIC

summary_output <- tibble(
  model_name = c("Foundational", "Health", "Qualifications", "Both"),
  aic = aics,
  bic = bics
)

summary_output |> 
  arrange(bic) |> 
  mutate(
    aic_rank = rank(aic),
    bic_rank = rank(bic)
  )

```

Lower BIC and AIC scores are better. The models are arranged from lowest to highest, and ranks shown as `aic_rank` and `bic_rank`.

This shows that the Qualifications model is preferred to the other three model specifications, then the Health model. The Foundation model is third, and the model with both appears too complex and is not preferred to the Foundational model.

There is no difference in the rank ordering of models preferred by the AIC and BIC metrics.

## Dilemma

-   The models with the best penalised fit are those considering **Qualifications** and **Health** separately

-   The model with the worst penalised fit is that with both **Qualifications** *and* **Health**.

-   But we are interested in something like *ceteris paribus* effects of intervening on one variable but not the other

We will present results from the Both model even though it has the poorest performance by these measures

## Representative data 

-   Select wave J from UKHLS as it's comparatively recent (and/but pre-pandemic)

```{r}
df_baseline <- ind_data_standardised |> 
  filter(wave == 'j')


df_counter_hlth <- 
  df_baseline |> 
  mutate(
    sf12mcs_dv = sf12mcs_dv + 2^-0.5,
    sf12pcs_dv = sf12pcs_dv + 2^-0.5
  )

df_counter_hlth <- 
  df_baseline |> 
  mutate(
    sf12mcs_dv = sf12mcs_dv + 2^-0.5,
    sf12pcs_dv = sf12pcs_dv + 2^-0.5
  )

df_counter_qual <- 
  df_baseline |> 
    mutate(qual_group = as.character(qual_group)) |> 
    mutate(qual_group = ifelse(qual_group == "Some", "Degree", qual_group)) |>
    mutate(qual_group = ifelse(qual_group == "None", "Some", qual_group)) |> 
    mutate(qual_group = ordered(qual_group, levels = c("None", "Some", "Degree")))

df_counter_both <- 
  df_baseline |> 
    mutate(qual_group = as.character(qual_group)) |> 
    mutate(qual_group = ifelse(qual_group == "Some", "Degree", qual_group)) |>
    mutate(qual_group = ifelse(qual_group == "None", "Some", qual_group)) |> 
    mutate(qual_group = ordered(qual_group, levels = c("None", "Some", "Degree"))) |> 
    mutate(
      sf12mcs_dv = sf12mcs_dv + 2^-0.5,
      sf12pcs_dv = sf12pcs_dv + 2^-0.5
    )

  
```

Now let's calculate the scenarios

```{r}
scen_base_hlth <- calculate_baseline_counterfactual_distribution(
  mod_full, d_base = df_baseline, d_counter = df_counter_hlth
)

scen_base_qual <- calculate_baseline_counterfactual_distribution(
  mod_full, d_base = df_baseline, d_counter = df_counter_qual
)

scen_base_both <- calculate_baseline_counterfactual_distribution(
  mod_full, d_base = df_baseline, d_counter = df_counter_both
)

scens_combined <- 
  scen_base_hlth |> 
    mutate(scenario = ifelse(scenario == "counterfactual", "counter_health", scenario)) |> 
    bind_rows(
      scen_base_qual |> 
    mutate(scenario = ifelse(scenario == "counterfactual", "counter_quals", scenario)) |> 
      filter(scenario != "baseline")
    ) |> 
    bind_rows(
      scen_base_both |> 
        mutate(scenario = ifelse(scenario == "counterfactual", "counter_both", scenario)) |> 
        filter(scenario != "baseline")
    )
  
```

## Relative change, plotted 

Now let's at least visualise this

```{r}
scens_combined |>
  filter(scenario != "baseline") |> 
  filter(output == "relative") |> 
  ggplot(aes(x = value, y = state, colour = scenario, shape = scenario)) + 
    geom_point() + 
  geom_vline(xintercept = 1)
  
```
## Relative change, as a table


```{r}
scens_combined |>
  filter(scenario != "baseline") |> 
  filter(output == "relative") |> 
  mutate(value = 100 * value) |> 
  mutate(value = round(value, 1)) |> 
  pivot_wider(names_from = scenario, values_from = value) |> 
  mutate(
    state = ordered(state, levels = c("Employed", "Unemployed", "Inactive student", "Inactive care", "Inactive long term sick", "Inactive retired", "Inactive other"))
  ) |> 
  arrange(state) |> 
  select(-output)
```
## Absolute changes, as a table

```{r}
scens_combined |> 
  filter(output == "absolute") |> 
  select(-output) |> 
  pivot_wider(names_from = "scenario", values_from = 'value')

```
## Absolute, as a graph

```{r}
scens_combined |> 
  filter(output == "absolute") |> 
  select(-output) |> 
  ggplot(aes(x = state, y = value, group = scenario, fill = scenario)) + 
  geom_col(position = "dodge") +
  coord_flip()
```

## Table of changes in proportions under each scenario 

Let's see how the distribution of persons in each state is estimated to change 

```{r}
scens_combined |> 
  filter(output == "absolute") |> 
  select(-output) |> 
  group_by(scenario) |> 
  mutate(share = value / sum(value)) |> 
  mutate(share = round(100 * share, 2)) |> 
  select(-value) |> 
  pivot_wider(names_from = 'scenario', values_from = 'share')
```

## Summary of tables

- Improving general health estimated to reduce size of inactive long-term sick population by more than increasing qualifications
- After health, the additional qualifications effect on inactive LT sick are estimated to be somewhat marginal
- Both improved health and improved qualifications are associated with higher proportions of the population who are students, and thus economically inactive for this reason. 
- Improved health is expected to increase the size of the employed population by around 2.4% 
