---
title: "Mean mental health and physical health scores, UKHLS, Waves a, b, i, j"
subtitle: "And changes in MH and PH from a-b and i-j for going from different economic inactivity states"
author:
  - "Jon Minton"
  - "Martin Taulbut"
format: 
  html:
    warning: false
    code-fold: true
    message: false
    code-summary: "Show R Code"
  docx: 
    warning: false
    echo: false
    message: false
editor: visual
prefer-html: true

---

# Aim

The aims are:

- Produce a table of mean mental health and physical health scores for waves a, b, i, and j

# Processing 

Load the data 

```{r}

devtools::load_all(here::here("R"))

base_dir_location <- "big_data/UKDA-6614-stata/stata/stata13_se/ukhls"
library(tidyverse)
library(nnet)
library(knitr)
library(kableExtra)

# Individual level attributes 

varnames <-  c(
  
  "jbstat", "dvage", "sex", # econ status; age; sex
  "sf12mcs_dv", "sf12pcs_dv" # SF12 MH, SF12 PH

  )

extract_what <- c(
  "labels", "values", "labels", 
  "values", "values"

  )


ind_data <- get_ind_level_vars_for_selected_waves(
  varnames = varnames, vartypes = extract_what
)

ind_data_processed<- 
  ind_data |> 
  rename(age = dvage) |> 
  mutate(across(c(age, sf12mcs_dv, sf12pcs_dv), function(x) ifelse(x < 0, NA, x))) %>%
  filter(between(age, 16, 64)) %>%
  filter(complete.cases(.))

```

```{r}
mean_mh_ph_scores <-
ind_data_processed |> 
  filter(wave %in% c("a", "j")) |> 
  group_by(wave, this_status) |> 
  summarise(mean_mh = mean(sf12mcs_dv, na.rm = TRUE), mean_ph = mean(sf12pcs_dv, na.rm = TRUE)) |>
  ungroup()

write.csv(mean_mh_ph_scores, here::here("support", "mean_mh_ph_scores.csv"))

mean_mh_ph_scores |> 
  # in a table, place wave above the mean mh and mean ph scores
  pivot_wider(names_from = wave, values_from = c(mean_mh, mean_ph)) |>
  kable() |>
  kable_styling() |> 
  add_header_above(c(" " = 1, "MH" = 2, "PH" = 2)) 
  # TO DO: rename header columns: status, mh, ph, mh, ph
  


```

Now want to look at, for pairs a-b, and i-j, the change in mean MH and PH scores for each change in status

The following is the observed mean scores in wave a by the paired from-to status known from wave b:
```{r}

mean_mh_ph_scores_by_transition_a_to_b <- 
  ind_data_processed |> 
    filter(wave %in% c('a')) |> 
    group_by(this_status, next_status) |> 
    summarise(mean_mh = mean(sf12mcs_dv, na.rm = TRUE), mean_ph = mean(sf12pcs_dv, na.rm = TRUE)) |>
    ungroup() 

write.csv(mean_mh_ph_scores_by_transition_a_to_b, here::here("support", "mean_mh_ph_scores_by_transition_a_to_b.csv"))

mean_mh_ph_scores_by_transition_a_to_b

```

And the same for waves i to j:

```{r}
mean_mh_ph_scores_by_transition_i_to_j <- 
  ind_data_processed |> 
    filter(wave %in% c('i')) |> 
    group_by(this_status, next_status) |> 
    summarise(mean_mh = mean(sf12mcs_dv, na.rm = TRUE), mean_ph = mean(sf12pcs_dv, na.rm = TRUE)) |>
    ungroup() 

write.csv(mean_mh_ph_scores_by_transition_i_to_j, here::here("support", "mean_mh_ph_scores_by_transition_i_to_j.csv"))

mean_mh_ph_scores_by_transition_i_to_j

```

To Do next:

- CHANGES IN HEALTH associated with changes in economic inactivity status

```{r}
ch_mh_ph_a_to_b <- 
  ind_data_processed |> 
    filter(wave %in% c("a", "b")) |> 
    select(pidp, wave, sf12mcs_dv, sf12pcs_dv) |> 
    pivot_wider(names_from = wave, values_from = c(sf12mcs_dv, sf12pcs_dv)) %>% 
    filter(complete.cases(.)) |> 
    mutate(
      change_mh = sf12mcs_dv_b - sf12mcs_dv_a,
      change_ph = sf12pcs_dv_b - sf12pcs_dv_a
    ) |> 
    select(pidp, change_mh, change_ph)

mean_ch_ph_mh_by_transition_a_to_b <-
  ind_data_processed |> 
    filter(wave %in% c("a", "b")) |> 
    select(pidp, this_status, next_status) |> 
    left_join(ch_mh_ph_a_to_b, by = "pidp") |> 
    group_by(this_status, next_status) |> 
    summarise(
      mean_change_mh = mean(change_mh, na.rm = TRUE),
      mean_change_ph = mean(change_ph, na.rm = TRUE),
      n_obs = n()
    ) |> 
    ungroup() 

write.csv(mean_ch_ph_mh_by_transition_a_to_b, here::here("support", "mean_ch_ph_mh_by_transition_a_to_b.csv"))

mean_ch_ph_mh_by_transition_a_to_b
```

And the same, but i to j

```{r}

ch_mh_ph_i_to_j <- 
  ind_data_processed |> 
    filter(wave %in% c("i", "j")) |> 
    select(pidp, wave, sf12mcs_dv, sf12pcs_dv) |> 
    pivot_wider(names_from = wave, values_from = c(sf12mcs_dv, sf12pcs_dv)) %>% 
    filter(complete.cases(.)) |> 
    mutate(
      change_mh = sf12mcs_dv_j - sf12mcs_dv_i,
      change_ph = sf12pcs_dv_j - sf12pcs_dv_i
    ) |> 
    select(pidp, change_mh, change_ph)

mean_ch_ph_mh_by_transition_i_to_j <-
  ind_data_processed |> 
    filter(wave %in% c("i", "j")) |> 
    select(pidp, this_status, next_status) |> 
    left_join(ch_mh_ph_i_to_j, by = "pidp") |> 
    group_by(this_status, next_status) |> 
    summarise(
      mean_change_mh = mean(change_mh, na.rm = TRUE),
      mean_change_ph = mean(change_ph, na.rm = TRUE),
      n_obs = n()
    ) |> 
    ungroup()

write.csv(mean_ch_ph_mh_by_transition_i_to_j, here::here("support", "mean_ch_ph_mh_by_transition_i_to_j.csv"))

mean_ch_ph_mh_by_transition_i_to_j


```
