---
echo: false
message: false
warning: false
---

The following model extended model specifications were considered and compared with the foundational model specification: 

- `mod_mh`: Mental health only 
- `mod_ph`: Physical health only 
- `mod_ph_mh`: Physical health and mental health: no interaction term
- `mod_phmh`: Physical health and mental health: interaction term

```{r}
devtools::load_all(here::here('R'))

library(tidyverse)
# library(haven)
# library(here)
library(nnet)

# base_dir_location <- "big_data/UKDA-6614-stata/stata/stata13_se/ukhls"
# indresp_files <- dir(here(base_dir_location), pattern = "[a-z]_indresp.dta", full.names = TRUE)

varnames <-  c(
  "jbstat", "dvage", "sex", "sf12mcs_dv", "sf12pcs_dv"
  )

vartypes <- c(
  "labels", "values", "labels", "values", "values"
  )

df_ind <- get_ind_level_vars_for_selected_waves(varnames = varnames, vartypes = vartypes, waves = letters[1:11])

# Clean the data 
df_ind_sf12_standardised <-
  df_ind |>
  # dvage uses negative values to indicate missing. The code below explicitly turns them all to missing values
    mutate(across(c(dvage, sf12mcs_dv, sf12pcs_dv), function(x) ifelse(x < 0, NA, x))) %>%
    filter(complete.cases(.)) |>
    mutate(across(c(sf12mcs_dv, sf12pcs_dv), standardise_scores)) |> 
  # This renames dvage to age
    rename(age = dvage) |>
    filter(between(age, 16, 64))  


```


```{r}
#| cache: true


mod_00 <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5),
    data = df_ind_sf12_standardised, 
    maxit = 200
  )

mod_mh <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + sf12mcs_dv,
    data = df_ind_sf12_standardised,
    maxit = 200  
  )

mod_ph <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + sf12pcs_dv,
    data = df_ind_sf12_standardised,
    maxit = 200
  )

mod_ph_mh <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + sf12pcs_dv + sf12mcs_dv,
    data = df_ind_sf12_standardised,
    maxit = 200
  )

mod_phmh <- 
  nnet::multinom(
    next_status ~ this_status * sex + splines::bs(age, 5) + sf12pcs_dv*sf12mcs_dv,
    data = df_ind_sf12_standardised,
    maxit = 200
)

```

The AICs and BICs of these model specifications are as follows:


```{r}
AIC(mod_00, mod_mh, mod_ph, mod_ph_mh, mod_phmh) |>
    as_tibble(rownames = "model") |>
    left_join(
        BIC(mod_00, mod_mh, mod_ph, mod_ph_mh, mod_phmh) |>
            as_tibble(rownames = "model")
    ) |>
    mutate(
        aic_rank = rank(AIC),
        bic_rank = rank(BIC)
    ) |>
    knitr::kable()


```

Here the results are unequivocal: both the AIC and BIC metrics indicate the model including both mental and physical health factors, and an interaction term between them, should be used. 
